---
title: "ABM Day 3"
subtitle: "Lecture 2: Temporal structure"
author: Lionel Hertzog
format:
  revealjs: 
    slide-number: true
    chalkboard: 
      buttons: false
    preview-links: auto
    css: styles.css
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
## load libraries
library(tidyverse)
library(inlabru)
library(rstanarm)
```

## Where we are

::: incremental
-   Introduction to Latent Gaussian Models and inlabru (earlier this morning)
-   Adding temporal structure in models (now)
-   Adding spatial structure in models (after lunch break)
-   Spatio-temporal structure (to end this day)
:::

## What we saw: Latent Gaussian Models


-  Observation parts
-  Latent fields
-  Hyperpameters prior definition

$$
\eta_i = \underbrace{\beta_0 + \beta_1 * X_i}_{\substack{\text{fixed} \\ \text{effects}}} + \underbrace{f_1(s_i)}_{\substack{\text{spatial} \\ \text{effect}}} + \underbrace{f_2(t_i)}_{\substack{\text{temporal} \\ \text{effect}}} + \underbrace{f_3(group_i)}_{\substack{\text{random} \\ \text{intercept}}} + ...
$$



## A motivating example

Unemployment rate in France:

```{r echo=FALSE}
dat <- read.csv("fr_unemploymeny.csv")

ggplot(dat, aes(x = time, 
                y = unemployment)) +
  geom_point(color = "#7D3C98") +
  labs(x = "Time",
       y = "Unemployment rate",
       title = "French unemployment rate",
       subtitle = "Rate per trimester") +
  theme_bw()
```

How would you model the temporal dynamic in unemployment rate?

## Option 1: lm

We could fit a simple linear trend to the data:

$$
unemployment_i = \beta_0 + \beta_1 * time_i + \epsilon_i;  \epsilon_i \sim \mathcal{N}(0, \sigma_{\epsilon}²)
$$
```{r echo=FALSE, message=FALSE}
ggplot(dat, aes(x = time, 
                y = unemployment)) +
  geom_point(color = "#7D3C98") +
  stat_smooth(method = "lm", color = "blue")
  labs(x = "Time",
       y = "Unemployment rate",
       title = "French unemployment rate",
       subtitle = "With lm fit") +
  theme_bw()
```
## Option 1: lm

Limits of a simple linear trend fit are:

-  assume *global* and *linear* effect of time
-  we could capture non-linearity by adding polynomials, but this is unstable
-  issues with i.i.d. assumption - correlation in residuals

```{r echo=FALSE}
m <- lm(unemployment ~ time, data = dat)
e <- resid(m)
acf(ts(e), ylab = "Correlation", main = "Autocorrelation of lm residuals")
```


## Option 2: mixed-effect models

We could also add time in a random-intercept model:

$$
unemployment = \beta_0 + u_{t(i)} + \epsilon_i; u_{t(i)} \sim \mathcal{N}(0, \sigma_{t}²)
$$
```{r echo=FALSE, message=FALSE, warning=FALSE}
dat$time_f <- factor(dat$time)
m_mix <- stan_lmer(unemployment ~ 1 + (1 | time_f), data = dat, refresh = 0)

## plot the results
ppred <- posterior_predict(m_mix)
pred <- plyr::adply(ppred, 2, function(x) data.frame(unemployment = mean(x),
                                             lci = quantile(x, probs = 0.05),
                                             uci = quantile(x, probs = 0.95)))
pred$time <- dat$time

ggplot(dat, aes(x = time, 
                y = unemployment)) +
  geom_point(color = "#7D3C98") +
  geom_ribbon(data = pred, aes(ymin = lci, ymax=uci),
              alpha = 0.25) +
  geom_line(data = pred, color = "blue") +
  labs(x = "Time",
       y = "Unemployment rate",
       title = "French unemployment rate",
       subtitle = "With mixed-model fit") +
  theme_bw()

```

## Option 2: issues

The mixed-effect model assumes that:

-  Time points are exchangeable, the model as no memory
-  The time deviates $u_5$ can be as close to $u_{10}$ as it is to $u_6$
-  However, a general assumption would be that two points close in time should be more similar than two points more distant


## Temporal structure in inlabru

Various temporal structure model are available via INLA:

|Model | Assumption| inlabru syntax |
|------|-----------|----------------|
| AR1 | Each value depends on the previous | f(time, model="ar1") |
| RW1 | Increments are iid Gaussian | f(time, model="rw1") |
| RW2 |Second differences are iid Gaussian | f(time, model="rw2") |
| Seasonal | Periodic structure of period $s$ | f(time, model="seasonal", season.length=s) |
|------|-----------|----------------|

## AR1 structure

A gaussian regression model with an AR1 structure is defined by:

$$
y_{i} \sim \mathcal{N}(\eta_i, \sigma_{\epsilon}²)
$$
$$
\eta_i = \beta_0 + f_{t_i}
$$
$$
f_{t} = \phi * f_{t-1} + \nu_t; \nu \sim \mathcal{N}(0, \sigma_{\nu}²) 
$$

The AR1 has two parameters: (i) $\phi$ the strength of the temporal autocorrelation - $\phi \approx 0$ close to mixed-effect model, $\phi \approx 1$ close to a random walk; (ii) $\sigma_{\nu}$ the noise around the autoregressive signal.

## AR1 model in action

Here is a AR1 model fitted to the unemployment data:

```{r echo=TRUE}
m_ar <- bru(unemployment ~ Intercept(1) + ar_field(time, model = "ar1"),
            data = dat, family = "gaussian",
            control.family = list(hyper = list(
      prec = list(param = c(10, 100))
    )))
```

```{r echo=FALSE}
# grab the preds from the ar model
pred <- predict(m_ar, dat, ~ Intercept + ar_field) 

ggplot(dat, aes(x = time, 
                y = unemployment)) +
  geom_point(color = "#7D3C98") +
  geom_ribbon(data = pred, aes(ymin = q0.025, ymax=q0.975),
              alpha = 0.25) +
  geom_line(data = pred, aes(y = mean), color = "blue") +
  labs(x = "Time",
       y = "Unemployment rate",
       title = "French unemployment rate",
       subtitle = "With AR1 fit") +
  theme_bw()

```

## RW models

RW1 is defined as:

$$
\Delta x_i = x_i - x_{i-1} \sim \mathcal{N}(0, \sigma²) 
$$
RW2 is defined as:

$$
\Delta²x_i = (x_i - x_{i-1}) - (x_{i-1} - x_{i-2}) \sim \mathcal{N}(0, \sigma²) 
$$
-  RW1 - *where I am* depend on where I was -> rough dynamic
-  RW2 - *my trend* depend on my past trend -> smooth
-  Both model type have one parameter $\sigma$, for RW1 large $\sigma$ means large jumps between time points, for RW2 large $\sigma$ means abrupt change in the trend. 

## RW model in action

```{r echo=TRUE}
m_rw1 <- bru(unemployment ~ Intercept(1) + rw_field(time, model = "rw1"),
            data = dat, family = "gaussian",
            control.family = list(hyper = list(
      prec = list(param = c(10, 100))
    )))

m_rw2 <- bru(unemployment ~ Intercept(1) + rw_field(time, model = "rw2"),
            data = dat, family = "gaussian",
            control.family = list(hyper = list(
      prec = list(param = c(10, 100))
    )))
```

```{r echo=FALSE}
# grab the preds from the ar model
pred_rw1 <- predict(m_rw1, dat, ~ Intercept + rw_field) 
pred_rw2 <- predict(m_rw2, dat, ~ Intercept + rw_field) 

ggplot(dat, aes(x = time, 
                y = unemployment)) +
  geom_point(color = "#7D3C98") +
  geom_ribbon(data = pred_rw1, aes(ymin = q0.025, ymax=q0.975),
              alpha = 0.25, fill = "blue") +
  geom_line(data = pred_rw1, aes(y = mean), color = "blue") +
   geom_ribbon(data = pred_rw2, aes(ymin = q0.025, ymax=q0.975),
              alpha = 0.25, fill = "orange") +
  geom_line(data = pred_rw2, aes(y = mean), color = "orange") +
  labs(x = "Time",
       y = "Unemployment rate",
       title = "French unemployment rate",
       subtitle = "With RW fit - blue: RW1, orange: RW2") +
  theme_bw()

```


## When to choose which

-  AR1 captures medium frequency autocorrelation in the data
-  RW2 fits low frequency trends in the data
-  RW1 is similar to AR1 but with a different parametrization

In inlabru we can combine these elements, for instance:

$$
\eta_i = \beta_0 + \underbrace{u_{t_i}}_{\text{rw2}} + \underbrace{a_{t_i}}_{\text{ar1}} + ... 
$$

## Coding time


