---
title: "ABM Day 3"
subtitle: "Lecture 3: Spatial structure"
author: Lionel Hertzog
format:
  revealjs:
    theme: white
    css: assets/custom.scss
    slide-number: true
    chalkboard: false
    incremental: true
    code-overflow: wrap
    fontsize: 2em
    title-slide-attributes:
      data-background-image: ""
    katex: TRUE
include-after-body: assets/logo-inject.html
execute:
  warning: false
  message: false
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
## load libraries
library(tidyverse)
library(inlabru)
library(rstanarm)
library(sf)
library(patchwork)
library(INLA)
library(fmesher)
```

## Where we are

::: incremental
-   Introduction to Latent Gaussian Models and inlabru (this morning)
-   Adding temporal structure in models (this morning)
-   Adding spatial structure in models (now)
-   Spatio-temporal structure (to end this day)
:::

## What we saw: Latent Gaussian Models


-  Observation parts
-  Latent fields
-  Hyperpameters prior definition

$$
\eta_i = \underbrace{\beta_0 + \beta_1 * X_i}_{\substack{\text{fixed} \\ \text{effects}}} + \underbrace{f_1(s_i)}_{\substack{\text{spatial} \\ \text{effect}}} + \underbrace{f_2(t_i)}_{\substack{\text{temporal} \\ \text{effect}}} + \underbrace{f_3(group_i)}_{\substack{\text{random} \\ \text{intercept}}} + ...
$$

## Different shades of spatial data

Spatial information can be encoded in various ways, for instance:

```{r echo=FALSE, message=FALSE, warning=FALSE}

## first example unemployment rate in french department
uu <- read.csv2("unemployment_dep.csv", dec = ".")
uus <- select(uu, Code, Libellé, T3_1999)

## add shapes
dep <- st_read("~/LIF/IFN_stuff/data/geodata/gadm41_FRA.gpkg",
               layer = "ADM_ADM_2", quiet = TRUE)

uus <- inner_join(uus, select(dep, CC_2, geom), by = c("Code" = "CC_2"))
names(uus)[3] <- "unemployment"


g1 <- ggplot() +
  geom_sf(data = uus, aes(fill = unemployment, geometry = geom)) +
  scale_fill_continuous(type = "viridis",
                        name = "Rate") +
  theme_void() +
  labs(title = "Unemployment rate in 1999")

## second example species richness in NFI plots
conn <- inventR::connect_db()
g3f <- tbl(conn, dbplyr::in_schema("inv_exp_nm", "g3flore"))
e1c <- tbl(conn, dbplyr::in_schema("inv_exp_nm", "e1coord"))
e2p <- tbl(conn, dbplyr::in_schema("inv_exp_nm", "e2point"))

g3f %>%
  count(npp) %>%
  inner_join(e1c, by = "npp") %>%
  inner_join(e2p, by = "npp") %>%
  filter(greco == "D") %>%
  select(npp, xl, yl, zp, n) %>%
  collect() -> fdat

fdat <- st_as_sf(fdat, coords = 2:3, crs = 27572)
fdat$n <- as.numeric(fdat$n)

g2 <- ggplot() +
  geom_sf(data = fdat, aes(color = log(n)), size = 0.25) +
  scale_color_continuous(type = "viridis",
                         name = "Species\nrichness") +
  theme_void() +
  labs(title = "Plant diversity in the Vosges")
  
wrap_plots(g1, g2, ncol = 2)
```

Do you see other ways?

## Non-spatial models fitted to spatial data

We can fit non-spatially aware models to the unemployment departmental data like:

$$
unemployment_i \sim \mathcal{N}(\eta_i, \sigma_{\epsilon}²)
$$
$$
\eta_i = \beta_0 + f(department_i)
$$
$$
f(department_i) \sim \mathcal{N}(0, \sigma_{dep}²)
$$
And in R:

```{r warning=FALSE, message=FALSE, echo=TRUE}
m_stan <- stan_lmer(unemployment ~ 1 + (1 | Code), data = uus,
                    refresh = 0)
```

## Non-spatial models

We can plot the model predictions:

```{r echo=FALSE}
uus$pred <- ranef(m_stan)$Code[,1] + fixef(m_stan)

g3 <- ggplot() +
   geom_sf(data = uus, aes(fill = pred, geometry = geom)) +
   scale_fill_continuous(type = "viridis",
                         name = "Rate") +
   theme_void() +
   labs(title = "Predicted unemployment rate",
        subtitle = "Mixed-effect model") 

wrap_plots(g1, g3, ncol = 2)
```

What do you think?

## Spatial models for areal data

Various model structure exists to consider the spatial structure from **discrete** spatial units.

All are based on an adjacency matrix between the units - whom is neighbors with whom:

```{r message=FALSE, warning=FALSE}
neb <- spdep::poly2nb(dep)
plot(neb, st_as_sfc(dep))
```

## Besag: the founding step

A first model on areal data is the Besag model:

$$
unemployment_i \sim \mathcal{N}(\eta_i, \sigma_{\epsilon}²)
$$


$$
\eta_i = \beta_0 + f(department_i)
$$

$$
f(department_i) = f_i | f_{-i} \sim \mathcal{N}(\frac{1}{n_i}\sum_{j \sim i} f_j, \frac{\sigma_{u}²}{n_i})
$$

where $n_i$ is the number of neighbours of area $i$ and $j \sim i$ means $j$ is a neighbour of $i$.

In plain language: the value of area $i$ is **the average of its neighbours**, with uncertainty that **decreases** as it has more neighbours.

## Besag in INLA(bru)

In inlabru such model can be fitted as follow:

```{r echo=TRUE, eval=FALSE}
m_besag <- bru(unemployment ~ Intercept(1) + besag_field(dep, model = "besag", graph = W))
```

Key properties:

-  Strong spatial pooling: the value of a spatial unit is pulled towards the one of its neighbours
-  One hyperparameter: $\sigma_u²$ that controls the variation between the spatial units.

## BYM: going a step further

The Besag-York-Mollié model adds an i.i.d. component on top of the Besag model:

$$
f(department_i) = \underbrace{u_i}_{\substack{\text{spatially} \\ \text{structured}}} + \underbrace{v_i}_{\text{unstructured}}
$$

$$
u_i = BESAG; v_i \sim \mathcal{N}(0, \sigma_v²)
$$

This separates two sources of variations between the spatial units:

-  A variation structured by the neighborhood structure of the units - spatially smooth
-  A variation that is unit-specific - pure overdispersion

It is similar to a BESAG + random-intercept setting.

## BYM in INLA(bru)

In inlabru we can fit such model with the following:

```{r echo=TRUE, eval=FALSE}
m_bym <- bru(unemployment ~ Intercept(1) + besag_field(dep, model = "bym", graph = W))
```

It has two hyperparameters:

-  $\sigma_u$: the amount of structured spatial variation
-  $\sigma_v$: the amount of spatial overdispersion

## An alternative BYM2

In practice it is recommended to work with a reparametrization of the BYM model:

$$
f(department_i) = \sigma (\sqrt{1-\rho}v_i^* + \sqrt{\rho}u_i^*)
$$

With $u_i^*$ a scaled BESAG and $v_i^* \sim \mathcal{N}(0, 1)$.

This model has two hyperparameters:

-   $\rho$: the share of the spatial variation that is structured by the neighborhood between the units, 0 pure i.i.d and 1 purely BESAG model.
-   $\sigma$: the amount of variation between the spatial units

## Comparison of options so far

| Model | Components | Key assumptions | Hyperparameters | Use when |
|-------|------------|-----------------|-----------------|----------|
| BESAG/ICAR| Spatial only | Smooth spatial variation | $\sigma²$ | Prior on smoothness |
| BYM | Spatial + iid | Spatial + overdispersion | $\sigma_u²$, $\sigma_v²$ | Overdispersion likely |
| BYM2 | Spatial + iid | Same as BYM with reparametrization | $\sigma²$, $\rho$ | reccommended option |

## Try out of BYM2

```{r echo=TRUE}
W <- spdep::nb2mat(neb, style = "B")
uus$id <- 1:nrow(uus)

m_bym2 <- bru(unemployment ~ Intercept(1) + bym_field(id, model = "bym2", graph = W),
              data = uus, control.family = list(hyper = list(prec = list(fixed = TRUE))))
```

I drop the estimation of the residual deviation from the model to avoid conflict between $\sigma_{\epsilon}²$ and $\sigma$.

```{r}
summary(m_bym2)
```

## Try out of BYM2

```{r echo=FALSE}
pred <- predict(m_bym2, uus, ~ Intercept + bym_field)

g4 <- ggplot() +
   geom_sf(data = pred, aes(fill = mean, geometry = geom)) +
   scale_fill_continuous(type = "viridis",
                         name = "Rate") +
   theme_void() +
   labs(title = "Predicted unemployment rate",
        subtitle = "BYM2 model") 

wrap_plots(g1, g4, ncol = 2)
```

Evidence of overfit ...

## From areal to continuous spatial fields


```{r}
g2
```

[In essence the spatial structures that we saw so far are equivalent to the hierarchical structure, they only consider discrete units, while with continuous spatial fields we can estimate a value at any point in space: $f(s_i)$]{style="font-size: 0.75em;"}

## About Gaussian Process

To estimate a continuous spatial field models often rely on Gaussian Process defined as:

$$
f(s_i) \sim \mathcal{N}(0, K(\theta, s_i, s_{i'}))
$$
With $K$ a suitable covariance function with parameter(s) $\theta$.

## About covariance function

Assume that we want to estimate a spatial field on 10 observed locations. The covariance matrix is a 10x10 matrix:


\begin{bmatrix}
\sigma²_{1} & \sigma_1\sigma_2 & \cdots & \sigma_1\sigma_{10} \\
\sigma_2\sigma_1 & \sigma_2² & \cdots & \sigma_2\sigma_{10} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma_{10}\sigma_1 & \sigma_{10}\sigma_2 & \cdots & \sigma_{10}²
\end{bmatrix}



The covariance function should fill in this matrix given some parameters and the distance between the locations:

$$
Cov(s_1, s_2) = K(\theta, d(s_1, s_2))
$$

## The Matern covariance function

A covariance function that is powerful to use for estimating spatial fields is the *Matern* function, it has two parameters:

-  $\rho$: the range parameter, how far the spatial correlation reach, large values means a fast decay of the correlation with the distance
-  $\sigma$: the amount of variation, large values indicate strongly varying spatial fields.

## Some matern plots - sigma

```{r}
build_matern <- function(n = 60,
                         theta1 = log(0.5),
                         theta2 = log(1.5)){
  # ── 2. Matérn Spatial Field ───────────────────────────────────────────────────
# Build a regular grid
grid_size <- n
coords <- expand.grid(
  x = seq(0, 1, length.out = grid_size),
  y = seq(0, 1, length.out = grid_size)
)
# Create SPDE object on the grid points as mesh
mesh <- inla.mesh.2d(
  loc      = coords,
  max.edge = c(0.08, 0.2),
  cutoff   = 0.02
)
# Build SPDE with Matérn covariance (smoothness nu = 1, i.e. alpha = 2)
spde <- inla.spde2.matern(mesh = mesh, alpha = 2)
# Sample one realisation from the Matérn field
Q <- inla.spde2.precision(spde, theta = c(theta1, theta2))
field_sample <- inla.qsample(n = 1, Q = Q)[, 1]
# Project mesh values onto the regular grid for plotting
A_proj <- inla.spde.make.A(mesh = mesh, loc = as.matrix(coords))
coords$z <- as.vector(A_proj %*% field_sample)
p_matern <- ggplot(coords, aes(x = x, y = y, fill = z)) +
  geom_raster() +
  scale_fill_distiller(palette = "RdYlBu", name = expression(f[1](s))) +
  coord_fixed() +
  labs(
    title = "",
    x = "",
    y = ""
  ) +
  theme_bw() +
  theme(panel.grid   = element_blank(),
    axis.ticks   = element_blank(),
    axis.text    = element_blank()
  )
return(p_matern)
}


# ── Combine with patchwork ────────────────────────────────────────────────────
combined <- wrap_plots(build_matern(),
                       build_matern(theta1 = 1),
                       build_matern(theta1 = 5),
                       nrow = 1) +
  plot_annotation(
    title    = "Effect of the sigma parameter on the Matérn field",
    theme    = theme(
      plot.title    = element_text(size = 14, face = "bold"))
  )

combined
```

## Some matern plots - rho

```{r}
combined <- wrap_plots(build_matern(),
                       build_matern(theta2 = 2.5),
                       build_matern(theta2 = 5),
                       nrow = 1) +
  plot_annotation(
    title    = "Effect of the rho parameter on the Matérn field",
    theme    = theme(
      plot.title    = element_text(size = 14, face = "bold"))
  )

combined
```

## The issue of space

-  As the amount of locations increases, so does the covariance matrix associated
-  I.e. for 1000 observations the covariance matrix is 1000x1000
-  So the naive approach of storing and estimating the Matern parameters directly is quickly computationally prohibitive

## The SPDE insight: from dense covariance to sparse precision

-  a Gaussian field with a Matérn covariance is the stationary solution to a specific Stochastic Partial Differential Equation (SPDE)
-  The solution to this SPDE on a triangulated mesh has a sparse precision matrix $Q$ rather than a dense covariance matrix $Σ$
-  Sparsity means most entries are exactly zero, and inference scales as $O(n^{3/2})$ instead of $O(n^3)$

## The mesh: discretising space

[-  To solve the SPDE, INLA/inlabru discretises the spatial domain using a constrained refined Delaunay triangulation (CRDТ): **the mesh**
-  The mesh is a triangular grid that extends a bit beyond the spatial domain of interest to avoid boundary issues.
-  The finer the mesh the more accurate the approximation, but at higher computational costs]{style="font-size: 0.65em;"}


```{r}
bnd <- fm_extensions(cbind(0, 0), convex = c(1, 1.5))

mesh1 <- fm_mesh_2d_inla(
  boundary = bnd,
  max.edge = c(0.1, 0.5)
)

mesh2 <- fm_mesh_2d_inla(
  boundary = bnd,
  max.edge = c(0.2, 0.5)
)

mesh3 <- fm_mesh_2d_inla(
  boundary = bnd,
  max.edge = c(0.5, 0.5)
)

g1 <- ggplot() +
  geom_fm(data = mesh1) +
  theme_bw()

g2 <- ggplot() +
  geom_fm(data = mesh2) +
  theme_bw()

g3 <- ggplot() +
  geom_fm(data = mesh3) +
  theme_bw()

g1 + g2 + g3


```

## Setting the SPDE

Once we have a mesh, we can define a SPDE associated to it:

```{r eval=FALSE}
spde <- inla.spde2.pcmatern(
  mesh  = mesh,
  prior.range = c(50, 0.05),   # P(rho < 50) = 0.05
  prior.sigma = c(1,  0.01)    # P(sigma > 1)    = 0.01
)
```

The two hyperparameters are:

| Hyperparameters | Interpretation | PC prior |
|-----------------|----------------|----------|
| $\rho$ (range)  | Distance at which correlation $\approx$ 0.13 | $P(\rho < \rho_0) = \alpha$ |
| $\sigma$ (variation)  | Amplitude of spatial variation | $P(\sigma > \sigma_0) = \alpha$ |


## Setting the PC prior in practice

-  `prior.range = c(50, 0.05)` says "I think it is unlikely (5% chance) that the range is shorter than 50 units", shorter range means more complex fields, infinite range is a flat field.
-  `prior.sigma = c(1, 0.01)` says "I think it is very unlikely (1% chance) that the spatial standard deviation exceeds 1" — on the scale of the linear predictor

-  Some helps:

```{r eval=FALSE, echo=TRUE}
# 1. What is the maximum distance in your study area?
max_dist <- max(dist(coordinates(dat)))

# 2. A range smaller than ~5% of the domain is probably too short
range_lower <- 0.05 * max_dist
prior.range  <- c(range_lower, 0.05)

# 3. What is the total SD of your response (on the link scale)?
total_sd    <- sd(dat$y_link)   # e.g. log(y) for Poisson
prior.sigma <- c(total_sd, 0.01)
```

## Fitting the model in inlabru

```{r echo=FALSE}
## set up the mesh
m_d <- fm_mesh_2d_inla(loc = st_coordinates(fdat),
                       max.edge = c(100, 1e4),
                       max.n.strict = 1e4)

## set up spde
spde_field <- inla.spde2.pcmatern(mesh = m_d, 
                                  prior.range = c(10000, 0.05),
                                  prior.sigma = c(0.6, 0.01))
```

```{r echo=TRUE}
m_matern <- bru(n ~ Intercept(1) + matern_field(geometry, model = spde_field),
                data = fdat,
                family = "poisson")

summary(m_matern)
```

## Plotting the spatial field

```{r echo=FALSE}
greco <- st_read("../../IFN_stuff/data/geodata/greco_2154.gpkg", quiet = TRUE)

grid_full <- st_make_grid(greco[greco$greco == "D",],
                          n = 100)
grid_crop <- st_intersection(grid_full,
                             greco[greco$greco == "D",])
grid_xy <- st_as_sf(st_centroid(grid_crop))
grid_xy$ID <- 1:nrow(grid_xy)
st_geometry(grid_xy) <- "geometry"

```

```{r echo=FALSE}
pred <- predict(m_matern, grid_xy, ~ Intercept + matern_field)

pred_grid <- st_drop_geometry(pred)
pred_grid <- cbind(pred_grid, grid_crop)
pred_grid <- st_as_sf(pred_grid)

ggplot() +
  geom_sf(data = pred_grid, aes(fill = mean), color = NA) +
  scale_fill_continuous(type = "viridis") +
  theme_void() +
  labs(fill = "Predictions",
       title = "Predicted species diversity")
```


## Wrapp up

-  Spatial models with discrete spatial units are based on the neighborhood structure between the units
-  It is recommended to use a *BYM2* model that provide an estimation of the proportion of the spatial variation that is structured by the neighborhood structure
-  Continuous spatial fields can be costly to estimate - large covariance matrix
-  The SPDE approach combined with a mesh allow for reliable and fast estimation of continuous spatial fields in INLA(bru)
