---
title: "Exercise: Introduction to the Bayesian Framework"
subtitle: "Likelihoods and Priors"
author: "Alexander Massey"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    number-sections: true
    code-fold: false
    code-copy: true
    code-overflow: wrap
    highlight-style: github
    theme: flatly
    fontsize: 1.05em
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

## Background

This exercise builds intuition for the three ingredients of Bayesian inference:

1. **The likelihood** — how probable is the data under different parameter values?
2. **The prior** — what do we believe before seeing the data?
3. **The posterior** — how do we update our beliefs after seeing the data?

We focus on the first two components here. A recurring theme is that choices which seem innocuous — a flat prior, an arbitrary starting value — can have substantial and sometimes surprising consequences.

---

## Part 1: Maximum Likelihood Estimation

Before placing priors, it helps to understand what the likelihood surface looks like on its own. The **maximum likelihood estimate (MLE)** is the parameter value that makes the observed data most probable under the assumed model. In simple cases it has a closed form; in general, we find it numerically.

### The Binomial likelihood: bird breeding

A field ecologist observes **3 successful breeding attempts out of 10** tries ($k = 3$, $N = 10$). We model the number of successes as $k \sim \text{Binomial}(N, p)$ and want to estimate the breeding success probability $p$.

The log-likelihood is:

$$\ell(p) = \log \binom{N}{k} + k \log p + (N - k) \log(1 - p)$$

The MLE is analytically $\hat{p} = k/N = 0.3$, but we will find it numerically using `optim()`.

```{r binomial-loglik}
library(ggplot2)

# Log-likelihood for Binomial(N = 10, p) with k = 3 successes
loglik <- function(p, k = 3, N = 10) {
  dbinom(x = k, size = N, prob = p, log = TRUE)
}

# Numerical MLE via optim() — minimise the NEGATIVE log-likelihood
fit <- optim(
  par    = 0.5,
  fn     = function(p) -loglik(p),
  method = "Brent",          # 1-D bounded optimiser
  lower  = 1e-6,
  upper  = 1 - 1e-6
)
mle <- fit$par

# Plot the log-likelihood surface
p_grid <- seq(0.001, 0.999, length.out = 500)
df <- data.frame(p = p_grid, ll = sapply(p_grid, loglik))

ggplot(df, aes(p, ll)) +
  geom_line(colour = "#2c3e50", linewidth = 1.2) +
  geom_vline(xintercept = mle, colour = "#e07b54",
             linewidth = 1, linetype = "dashed") +
  annotate("text", x = mle + 0.05, y = min(df$ll) + 0.8,
           label = paste0("MLE = ", round(mle, 3)),
           colour = "#e07b54", size = 5) +
  labs(x = "p", y = "Log-likelihood",
       title = "Binomial log-likelihood:  k = 3,  N = 10") +
  theme_minimal(base_size = 14)
```

**Questions:**

a. Change the data to $k = 0$ successes. What is the MLE? Plot the log-likelihood. What does this imply about using the MLE as a point estimate in small samples with extreme outcomes — and how does a Bayesian prior help?
b. The log-likelihood is quite flat over a wide range of $p$. What does this tell you about the **information content** of $N = 10$ observations?  What happens if we increase $N$ to 100 (keeping $\hat{p} = 0.3$, so $k = 30$) and replot. How does the surface change?

::: {.callout-tip collapse="true"}
## Hint 1a — Zero successes
```r
loglik_0 <- function(p) dbinom(x = 0, size = 10, prob = p, log = TRUE)
fit_0 <- optim(par = 0.5, fn = function(p) -loglik_0(p),
               method = "Brent", lower = 1e-6, upper = 1 - 1e-6)
fit_0$par   # → near 0
```
The MLE is $\hat{p} = 0$ — the model assigns zero probability to ever observing a success. A Beta(1,1) or weakly informative prior immediately regularises this toward the interior of $[0,1]$.
:::

::: {.callout-tip collapse="true"}
## Hint 1b — Larger sample
```r
loglik_large <- function(p) dbinom(x = 30, size = 100, prob = p, log = T)
loglik_small <- function(p) dbinom(x = 3, size = 10, prob = p, log = T)
p_grid <- seq(0.001, 0.999, length.out = 500)
df_large <- data.frame(p = p_grid, ll = sapply(p_grid, loglik_large))
df_small <- data.frame(p = p_grid, ll = sapply(p_grid, loglik_small))
ggplot(df_large, aes(p, ll)) + geom_line(color = "blue") +
 geom_line(data = df_small, color = "red") + theme_minimal()
```
The surface becomes much sharper (more peaked) around 0.3 — the data are more informative, leaving less posterior uncertainty regardless of the prior.
:::

---

### MLE for a Normal distribution

We now work with a two-parameter model. Suppose we have collected heights (in cm) from 100 individuals. We assume $\text{height}_i \sim \mathcal{N}(\mu, \sigma^2)$ and want to find the MLEs of $\mu$ and $\sigma$ numerically.

The log-likelihood is:

$$\ell(\mu, \sigma) = -\frac{n}{2}\log(2\pi\sigma^2) - \frac{1}{2\sigma^2}\sum_{i=1}^n (x_i - \mu)^2$$

The MLEs are analytically $\hat{\mu} = \bar{x}$ and $\hat{\sigma} = \sqrt{\frac{1}{n}\sum(x_i - \bar{x})^2}$, but `optim()` finds them in the same framework we will use for models without closed forms.

```{r normal-mle}
set.seed(2020)
n      <- 100
height <- rnorm(n, mean = 170, sd = 10)

# Negative log-likelihood for Normal(mu, sigma)
negloglik <- function(theta, data) {
  mu    <- theta[1]
  sigma <- theta[2]
  -sum(dnorm(data, mean = mu, sd = sigma, log = TRUE))
}

# Check at an arbitrary starting point
negloglik(theta = c(150, 1), height)

# Optimise jointly over (mu, sigma)
fit <- optim(
  par = c(mean(height), sd(height)),   # sensible starting values
  fn  = negloglik,
  data = height
)
fit$par   # should recover ~ (170, 10)
```

**Questions:**

a. Run `optim()` with the deliberately bad starting values `par = c(150, 1)`. Does it still converge to the correct MLEs? Now try `par = c(0, 0.001)`. What goes wrong, and why?
b. Compare `fit$par` to `c(mean(height), sd(height))`. Are they identical? The MLE for $\sigma$ is the *biased* population standard deviation $\hat{\sigma} = \sqrt{\frac{1}{n}\sum(x_i - \bar{x})^2}$, not R's `sd()` which uses $n-1$. Is the difference meaningful here at $n = 100$?

::: {.callout-tip collapse="true"}
## Hint 2a — Bad starting values
`sigma = 0.001` causes `dnorm()` to return `-Inf` for most observations (essentially assigning zero density to data far from the mean). `optim()` cannot compute a gradient and fails. This illustrates that numerical optimisers are sensitive to the scale of the parameter space — a practical reason to work on the log scale for variance parameters.
:::

::: {.callout-tip collapse="true"}
## Hint 2b
The MLE is consistent but not guaranteed to be unbiased. Using $n-1$ corrects this bias so it should be larger.
:::


---

## Part 2: The Pitfalls of Flat Priors

A **flat (uniform) prior** over a parameter expresses "complete ignorance" — every value is equally plausible before seeing data. This sounds reasonable but causes two related problems:

1. **It is not invariant to reparameterisation.** A flat prior on $p$ is *not* flat on $\log(p/(1-p))$ — the model "knows" more than we think it does.
2. **It might place substantial mass in implausible regions**, especially on unbounded or transformed scales.

### Flat priors on a probability scale

Consider the logistic regression model for binary survival $\phi_i$:

$$\text{logit}(\phi_i) = \beta_0 + \beta_1 x_i, \qquad \beta_0, \beta_1 \sim \mathcal{N}(0, \tau^2)$$

The prior on $\phi_i = \text{logit}^{-1}(\beta_0 + \beta_1 x_i)$ depends critically on $\tau$.

```{r flat-prior-logistic}
set.seed(42)
n_sim <- 10000

# "Flat" prior on the logit scale — large variance
beta_flat <- rnorm(n_sim, 0, 10)
phi_flat  <- plogis(beta_flat)   # back-transform to probability scale

# Weakly informative prior
beta_weak <- rnorm(n_sim, 0, 1.5)
phi_weak  <- plogis(beta_weak)

par(mfrow = c(1, 2))
plot(density(phi_flat, from = 0, to = 1),
     main = expression(beta ~ "~ N(0, 10²)"),
     xlab = "Survival probability", ylim = c(0, 8))
plot(density(phi_weak, from = 0, to = 1),
     main = expression(beta ~ "~ N(0, 1.5²)"),
     xlab = "Survival probability", ylim = c(0, 8))
par(mfrow = c(1, 1))
```

**Questions:**

a. Describe the shape of the implied prior on $\phi$ when $\beta \sim \mathcal{N}(0, 10^2)$. Is this really "non-informative"? What claim does this prior make?
b. How does the implied prior change with $\mathcal{N}(0, 1.5^2)$? Is this more or less informative in the logit space? Why is it *more appropriate* despite being "informative"?
c. Try $\tau = 0.5$ and $\tau = 5$. Describe the trade-off. Is there a value of $\tau$ you would defend for a survival probability in ecology?

::: {.callout-tip collapse="true"}
## Hint 3a
$\mathcal{N}(0, 10^2)$ on the logit scale concentrates mass near 0 and 1 on the probability scale — it is actually *highly* informative, favouring extreme survival probabilities. This is the opposite of what "non-informative" should mean for a bounded probability.
:::

::: {.callout-tip collapse="true"}
## Hint 3b
$\mathcal{N}(0, 1.5^2)$ on the logit scale maps to an approximately uniform distribution on the probability scale — it is diffuse where it matters (on the constrained $[0,1]$ space) and avoids the pathological spike at the boundaries. Informativeness is always relative to the parameterisation.
:::

### Flat priors in linear regression

A flat prior on regression coefficients — $\beta \sim \mathcal{N}(0, 100^2)$ — is similarly deceptive when predictors are on a natural scale.  Consider a standardized predictor's influence on human height.

```{r flat-prior-regression}
set.seed(42)
n_sim <- 10000

# "Flat" prior: beta ~ N(0, 100^2) on a standardised predictor
beta_flat_lm <- rnorm(n_sim, 0, 100)

# Weakly informative prior: beta ~ N(0, 1^2)
beta_weak_lm <- rnorm(n_sim, 2, 0.5)

par(mfrow = c(1, 2))
hist(beta_flat_lm, breaks = 60, main = expression(beta ~ "~ N(0, 100²)"),
     xlab = expression(beta), freq = FALSE, col = "#b0c4de")
hist(beta_weak_lm, breaks = 60, main = expression(beta ~ "~ N(0, 1²)"),
     xlab = expression(beta), freq = FALSE, col = "#90c4a8")
par(mfrow = c(1, 1))
```

**Questions:**

a. Why is the "flat" prior unreasonable?


::: {.callout-tip collapse="true"}
## Hint 4a
It could easily produce negative heights!  You would really have to trust your data to not have outliers or measurement error because those impossible values are permissible under the prior. 
:::

---

## Part 3: Eliciting a Prior via Method of Moments

When external information is available — from published studies, expert knowledge, or pilot data — we should incorporate it into the prior rather than defaulting to flatness. **Method of moments** is a simple, transparent way to translate summary statistics (mean and variance) into the parameters of a prior distribution.

### The Beta distribution and moment matching

We are modelling a detection probability $p$ in a capture-recapture study. From the literature, we believe:

- **Prior mean:** $\mu = 0.57$ (detection is reasonably high)
- **Prior variance:** $\sigma^2 = 0.073^2 = 0.005329$ (moderate uncertainty)

We want a $\text{Beta}(\alpha, \beta)$ prior consistent with these moments. Recall that for the Beta distribution:

$$\mu = \frac{\alpha}{\alpha + \beta}, \qquad \sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$$

**Questions:**

a. Solve analytically for $\alpha$ and $\beta$ in terms of $\mu$ and $\sigma^2$. Show your algebra — express $\alpha + \beta$ first (sometimes called the *concentration* or *precision* $\kappa$), then recover $\alpha$ and $\beta$.
b. Implement your solution in R and compute the numerical values of $\alpha$ and $\beta$.
c. Verify your answer by simulation: draw 50 000 samples from $\text{Beta}(\hat{\alpha}, \hat{\beta})$ and check that the sample mean and variance match the target moments.

::: {.callout-tip collapse="true"}
## Hint 5a — Algebra
From the mean equation: $\alpha = \mu(\alpha + \beta)$, so $\beta = \alpha(1-\mu)/\mu$.

Substitute into the variance equation. After simplification:

$$\alpha + \beta = \frac{\mu(1-\mu)}{\sigma^2} - 1$$

Let $\kappa = \alpha + \beta$. Then $\alpha = \kappa\mu$ and $\beta = \kappa(1 - \mu)$.
:::

::: {.callout-tip collapse="true"}
## Hint 5b — R implementation
```r
mu    <- 0.57
sigma2 <- 0.073^2

kappa <- mu * (1 - mu) / sigma2 - 1
alpha <- kappa * mu
beta  <- kappa * (1 - mu)

cat("alpha =", round(alpha, 2), "\nbeta  =", round(beta,  2))
```
:::

::: {.callout-tip collapse="true"}
## Hint 5c — Simulation check
```r
set.seed(123)
sims <- rbeta(50000, shape1 = alpha, shape2 = beta)

cat("Simulated mean:", round(mean(sims), 4),   "  Target:", mu,     "\n")
cat("Simulated var: ", round(var(sims),  6),   "  Target:", sigma2, "\n")

# Visualise
plot(density(sims), main = "Beta prior via moment matching",
     xlab = "Detection probability p")
abline(v = mu, col = "firebrick", lty = 2, lwd = 2)
```
:::

### Sensitivity to the prior

Once you have an informed prior, it is good practice to check how sensitive your conclusions are to reasonable perturbations of it.

**Questions:**

a. Construct two alternative priors: one **more diffuse** ($\sigma^2 = 0.15^2$) and one **more concentrated** ($\sigma^2 = 0.03^2$), both centred at $\mu = 0.57$. Compute and plot all three Beta priors on the same axes.
b. Suppose you observe $k = 8$ detections in $N = 20$ trials. The posterior is $\text{Beta}(\alpha + k,\; \beta + N - k)$. Compute and plot the three posteriors. How much does the choice of prior matter here?

::: {.callout-tip collapse="true"}
## Hint 6a — Three priors
```r
mu <- 0.57;  k <- 8;  N <- 20

make_beta_params <- function(mu, sigma2) {
  kappa <- mu * (1 - mu) / sigma2 - 1
  c(alpha = kappa * mu, beta = kappa * (1 - mu))
}

p1 <- make_beta_params(mu, 0.073^2)   # informed
p2 <- make_beta_params(mu, 0.15^2)    # diffuse
p3 <- make_beta_params(mu, 0.03^2)    # concentrated

p_grid <- seq(0.001, 0.999, length.out = 500)
plot(p_grid, dbeta(p_grid, p1["alpha"], p1["beta"]),
     type = "l", col = "#2c7bb6", lwd = 2,
     xlab = "p", ylab = "Density", main = "Three Beta priors")
lines(p_grid, dbeta(p_grid, p2["alpha"], p2["beta"]), col = "#d7191c", lwd = 2)
lines(p_grid, dbeta(p_grid, p3["alpha"], p3["beta"]), col = "#1a9641", lwd = 2)
legend("topleft", legend = c("Informed", "Diffuse", "Concentrated"),
       col = c("#2c7bb6", "#d7191c", "#1a9641"), lwd = 2)
```
:::

::: {.callout-tip collapse="true"}
## Hint 6b — Posteriors
```r
# Posterior = Beta(alpha + k, beta + N - k)
plot(p_grid, dbeta(p_grid, p1["alpha"] + k, p1["beta"] + N - k),
     type = "l", col = "#2c7bb6", lwd = 2,
     xlab = "p", ylab = "Density", main = "Posteriors after 8/20 detections")
lines(p_grid, dbeta(p_grid, p2["alpha"] + k, p2["beta"] + N - k),
      col = "#d7191c", lwd = 2)
lines(p_grid, dbeta(p_grid, p3["alpha"] + k, p3["beta"] + N - k),
      col = "#1a9641", lwd = 2)
legend("topleft", legend = c("Informed", "Diffuse", "Concentrated"),
       col = c("#2c7bb6", "#d7191c", "#1a9641"), lwd = 2)
```
With $N = 20$ observations, the data dominate and the three posteriors should be nearly identical. Repeat with $N = 5$ to see prior sensitivity increase.
:::

---

## References

Gimenez, O. (2025). *Bayesian statistics with R*. Lecture slides, November–December 2025. <https://oliviergimenez.github.io/bayesian-stats-with-R/>
