---
title: "Multi-level models"
subtitle: "tadpole example from Rethinking"
author: "Alexander Massey"
date: today
date-format: "MMMM D, YYYY"
format:
  revealjs:
    embed-resources: true
    theme: white
    slide-number: true
    chalkboard: false
    incremental: true
    code-overflow: wrap
    fontsize: 2em
    title-slide-attributes:
      data-background-image: ""
    katex: TRUE
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
```

## Outline {.unnumbered}

::: {.nonincremental}
1. The problem with independent group estimates
2. What is a multilevel model?
3. Mixed models: same idea, different language
4. Three estimators: complete pooling, no pooling, partial pooling
5. Bayesian hierarchical specification
6. Shrinkage and why it works
7. Today's exercise
:::


# Part 1 {.section-slide}

## The Core Problem

Suppose you observe survival rates across **48 tanks** of tadpoles.

. . .

Each tank has $n_i$ tadpoles and $s_i$ survivors.

. . .

**Two naive strategies:**

- Estimate each tank independently → noisy, overfits small tanks
- Pool all tanks together → ignores real variation between tanks

. . .

::: {.callout-note appearance="minimal"}
Neither extreme is satisfying. We need something in between.
:::

---

## A Concrete Example

::: {.columns}
::: {.column width="50%"}
**Tank A:** 5 tadpoles, 5 survive  
$\hat{p} = 1.00$

**Tank B:** 5 tadpoles, 0 survive  
$\hat{p} = 0.00$
:::
::: {.column width="50%"}
Should we really believe survival is **100%** or **0%** in these tanks?

. . .

Or is there **sampling noise** on top of a shared population process?
:::
:::

. . .

The key insight: tanks are not isolated experiments — they share a common biology.


# Part 2 {.section-slide}

## What Is a Multilevel Model?

A multilevel (hierarchical) model adds a **population distribution** over group-level parameters.

. . .

Instead of:

$$\alpha_{\text{tank}} \sim \text{Normal}(0,\ 5) \quad \text{(independent, fixed prior)}$$

. . .

We write:

$$\alpha_{\text{tank}} \sim \text{Normal}(\mu_\alpha,\ \sigma_\alpha) \quad \text{(shared population)}$$

$$\mu_\alpha \sim \text{Normal}(0, 1), \qquad \sigma_\alpha \sim \text{Half-Cauchy}(0, 1)$$

. . .

Now $\mu_\alpha$ and $\sigma_\alpha$ are **estimated from the data**.

---

## What This Buys Us

- **Regularization**: small tanks are pulled toward the population mean
- **Calibrated uncertainty**: wide $\sigma_\alpha$ → tanks really differ; small $\sigma_\alpha$ → pooling is justified
- **Honest inference**: we acknowledge that groups come from a common population

. . .

This is not a trick — it is the **correct generative model** when groups share a common process.


# Part 3 {.section-slide}

## Mixed Models: Same Idea, Different Language

Frequentist **linear mixed models** (LMMs) are hierarchical models in disguise.

. . .

In `lme4`:

```r
lmer(surv ~ 1 + (1 | tank), data = d)
```

This fits the same structure as our Bayesian model — a population-level intercept plus tank-level deviations.

. . .

So why the confusion?

---

## The Terminology Problem

::: {style="font-size: 0.85em;"}

| Frequentist LMM term | What it actually means |
|---|---|
| **Fixed effect** | Population-level parameter (estimated, interpreted directly) |
| **Random effect** | Group-level deviation from the population mean |
| **REML** | Restricted maximum likelihood — a bias correction for variance components; not a prior |
| **BLUPs** | Best linear unbiased predictors — empirical Bayes estimates of group-level effects |

:::

. . .

::: {.callout-warning appearance="minimal"}
"Fixed" and "random" do not mean what they sound like. The terminology predates modern statistical thinking and causes genuine confusion even among practitioners.
:::

---

## Bayesian vs. Frequentist Framing

| | Bayesian (brms) | Frequentist (lme4) |
|---|---|---|
| Group intercepts | Parameters with a prior | "Random effects" — not full parameters |
| Population $\sigma$ | Posterior distribution | Point estimate via REML |
| Uncertainty | Full posterior | Confidence intervals (often approximate) |
| Interpretation | Direct probability statements | Repeated-sampling guarantees |

. . .

The **models are nearly identical**. The Bayesian framing is more explicit about what is being assumed.  Also the prior helps with toxic cases like if all units in group are identical.


# Part 4 {.section-slide}

## Three strategies:

**Complete pooling** — All tanks share a single intercept
$$\text{logit}(p_i) = \alpha$$

**No pooling** — estimate each group independently
$$\text{logit}(p_i) = \alpha_{\text{tank}[i]}$$

**Partial Pooling** — share population distribution so information is "borrowed" across groups (shrinkage).

$$\begin{aligned} \text{logit}(p_i) &= \alpha_{\text{tank}[i]} \\ \alpha_{\text{tank}} &\sim \text{Normal}(\mu_{\alpha}, \sigma_{\alpha}) \\ \mu_{\alpha} &\sim \text{Normal}(0, 1.5) \\ \sigma_{\alpha} &\sim \text{Exponential}(1) \end{aligned}$$

---

## Three strategies:

**Complete pooling** 

```r
# All tanks share a single intercept
brm(surv | trials(density) ~ 1, 
    family = binomial, data = d)
```

**No pooling** 
```r
brm(surv | trials(density) ~ 0 + tank, 
    family = binomial, data = d)
```

**Partial Pooling** 

```r
# Adaptive/Varying intercepts for tank
brm(surv | trials(density) ~ 1 + (1 | tank),
    data = d, family = binomial,
    prior = c(
      set_prior("normal(0, 1.5)", class = "Intercept"), # the default
      set_prior("exponential(1)", class = "sd")         # the default
    ))
```

---

## The Trade-off

```{r}
#| label: pooling-diagram
#| fig-height: 3.5
#| fig-width: 9

set.seed(42)
n <- 20
true_mu <- 0.6
true_sd <- 0.25
true_p  <- plogis(rnorm(n, qlogis(true_mu), true_sd))
ni      <- sample(c(3, 5, 8, 20, 30), n, replace = TRUE)
si      <- rbinom(n, ni, true_p)
p_obs   <- si / ni

# Partial pool: shrink toward grand mean, weighted by ni
grand   <- weighted.mean(p_obs, ni)
lambda  <- ni / (ni + 5)
p_part  <- lambda * p_obs + (1 - lambda) * grand

df <- tibble(
  tank   = 1:n,
  nopool = p_obs,
  partial= p_part,
  complete = grand
)

df %>%
  pivot_longer(-tank, names_to = "estimator", values_to = "p") %>%
  mutate(estimator = factor(estimator,
    levels = c("nopool", "partial", "complete"),
    labels = c("No pooling", "Partial pooling", "Complete pooling"))) %>%
  ggplot(aes(x = tank, y = p, color = estimator, shape = estimator)) +
  geom_hline(yintercept = true_mu, linetype = "dashed", color = "grey60", linewidth = 0.5) +
  geom_point(size = 2.5, alpha = 0.9) +
  scale_color_manual(values = c("No pooling" = "#E07B39",
                                "Partial pooling" = "#2C6E8A",
                                "Complete pooling" = "#555555")) +
  scale_shape_manual(values = c("No pooling" = 16, "Partial pooling" = 1, "Complete pooling" = 3)) +
  scale_y_continuous(limits = c(0, 1), name = "Estimated survival probability") +
  scale_x_continuous(name = "Tank") +
  annotate("text", x = n + 0.5, y = true_mu + 0.03, label = "true μ",
           size = 3, color = "grey50", hjust = 0) +
  labs(color = NULL, shape = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())
```

. . .

Partial pooling is **adaptive**: small tanks shrink more, large tanks shrink less.

# Part 5 {.section-slide}

## Specify full hierarchical Bayesian model:

**Level 1** — observation model (likelihood):

$$s_i \sim \text{Binomial}(n_i,\ p_i)$$


**Level 2** — group-level model:

$$\text{logit}(p_i) = \alpha_{\text{tank}_i}$$
$$\alpha_{\text{tank}} \sim \text{Normal}(\mu_\alpha,\ \sigma_\alpha)$$

**Level 3** — hyperpriors:

$$\mu_\alpha \sim \text{Normal}(0,\ 1)$$
$$\sigma_\alpha \sim \text{Half-Cauchy}(0,\ 1)$$

---

## In brms

**No-pooling model** — each tank gets an independent intercept:

```r
brm(surv | trials(density) ~ 0 + factor(tank),
    prior = prior(normal(0, 5), class = b),
    family = binomial, data = d)
```

. . .

**Partial-pooling model** — intercepts drawn from a population:

```r
brm(surv | trials(density) ~ 1 + (1 | tank),
    prior = c(prior(normal(0, 1), class = Intercept),
              prior(cauchy(0, 1), class = sd)),
    family = binomial, data = d)
```

. . .

The syntax `(1 | tank)` is borrowed from `lme4`. The prior on `class = sd` controls $\sigma_\alpha$ — **in the standard deviation metric**.

---

## Priors on the Variance Component

$\sigma_\alpha$ governs how much tanks vary. Two common choices:

. . .

**Half-Cauchy(0, 1)** — heavy tails, allows large between-group variation

$$\sigma_\alpha \sim \text{Half-Cauchy}(0,\ 1)$$

. . .

**Exponential(1)** — lighter tails, more skeptical of extreme variation

$$\sigma_\alpha \sim \text{Exponential}(1)$$

. . .

In practice, for well-identified problems with moderate group counts, the choice rarely matters much. **Check with a sensitivity analysis.**

# Part 6 {.section-slide}

## Shrinkage: Why It Works

When a tank has **few observations**, the likelihood is weak.

. . .

The population prior $\text{Normal}(\mu_\alpha, \sigma_\alpha)$ dominates → estimate pulled toward $\mu_\alpha$.

. . .

When a tank has **many observations**, the likelihood is strong.

. . .

The data dominate → estimate close to the observed proportion.

. . .

::: {.callout-note appearance="minimal"}
This is **automatic regularization** — no tuning parameter required. The degree of shrinkage is determined by the data through the estimated $\sigma_\alpha$.
:::

---

## Shrinkage Reduces Error

```{r}
#| label: error-comparison
#| fig-height: 3.2
#| fig-width: 9

set.seed(7)
ns    <- c(3, 5, 10, 25, 40)
reps  <- 200
true_mu_lo <- 0.6
true_sd_lo <- 0.3

results <- map_dfr(ns, function(n_i) {
  map_dfr(1:reps, function(r) {
    true_p  <- plogis(rnorm(1, qlogis(true_mu_lo), true_sd_lo))
    s       <- rbinom(1, n_i, true_p)
    p_nopool <- s / n_i
    # simple shrinkage toward 0.6
    lambda  <- n_i / (n_i + 8)
    p_part  <- lambda * p_nopool + (1 - lambda) * true_mu_lo
    tibble(n = n_i,
           nopool_err  = abs(p_nopool - true_p),
           partial_err = abs(p_part   - true_p))
  })
})

results %>%
  group_by(n) %>%
  summarise(nopool  = mean(nopool_err),
            partial = mean(partial_err)) %>%
  pivot_longer(-n, names_to = "estimator", values_to = "mean_abs_error") %>%
  mutate(estimator = if_else(estimator == "nopool", "No pooling", "Partial pooling")) %>%
  ggplot(aes(x = n, y = mean_abs_error, color = estimator)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_color_manual(values = c("No pooling" = "#E07B39", "Partial pooling" = "#2C6E8A")) +
  scale_x_continuous(breaks = ns, name = "Tadpoles per tank") +
  scale_y_continuous(name = "Mean absolute error") +
  labs(color = NULL,
       title = "Partial pooling beats no pooling most in small groups") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())
```

. . .

The advantage is largest for **small groups** and shrinks as within-group sample size grows.

---

## When Is Partial Pooling Worth It?

Partial pooling is most valuable when:

- Some groups have **few observations**
- Groups plausibly share a **common population process**
- You want to compensate for **compounding variables** implicitly
- You plan to **predict for new groups** not seen in training

. . .

It converges to complete pooling when $\hat{\sigma}_\alpha \to 0$, and to no pooling when $\hat{\sigma}_\alpha \to \infty$.


# Part 7 {.section-slide}

## Today's Exercise

**Dataset:** Reed frog tadpole survival (`reedfrogs`, 48 tanks)

. . .

**Part 1 — Real data**

- Fit the no-pooling and partial-pooling binomial models
- Compare predictive accuracy with `loo_compare()`
- Visualize shrinkage of tank estimates

. . .

**Part 2 — Simulation study**

- Simulate 60 ponds with known true survival probabilities
- Compare all three estimators against ground truth
- Confirm that partial pooling dominates in small ponds

---

## Workflow for Each Model

For every model you fit today:

::: {.nonincremental}
1. **Prior predictive check** — do prior draws look plausible?
2. **Fit the model** — `brm(...)`
3. **Convergence** — inspect $\hat{R}$, $N_{eff}$, trace plots
4. **Posterior predictive check** — does the model reproduce the data?
5. **Inference** — interpret parameters, plot effects
:::

. . .

This workflow applies to every Bayesian model, not just multilevel ones. Make it a habit.

---

## Key Parameters to Interpret

From `print(model_brms)`:

- `b_Intercept` → $\mu_\alpha$, population-average log-odds of survival
  → back-transform: `plogis(b_Intercept)`

- `sd_tank__Intercept` → $\sigma_\alpha$, between-tank standard deviation on log-odds scale

. . .

From `coef(model_brms)$tank`:

- Tank-level posterior medians, partially pooled toward $\mu_\alpha$

. . .

From `loo_compare()`:

- `elpd_diff` — difference in expected log predictive density
- Rule of thumb: `|elpd_diff| > 2 × se_diff` → meaningful difference

---

## Summary

::: {.nonincremental}
- Multilevel models share information across groups via a **population distribution**
- They are identical in structure to frequentist mixed models — the Bayesian framing is more explicit
- Frequentist terminology (fixed/random effects, REML) is a historical artifact; the Bayesian terms are clearer
- Partial pooling **adaptively regularizes**: more shrinkage for data-sparse groups
- In `brms`: `(1 | group)` specifies varying intercepts; prior on `class = sd` controls $\sigma_\alpha$
:::

. . .

::: {.callout-note appearance="minimal"}
Open `multilevel_exercise.qmd`.
:::

---

## References

::: {.nonincremental style="font-size: 0.8em;"}
- McElreath, R. (2020). *Statistical Rethinking*, 2nd ed. CRC Press. Chapter 13.
- Kurz, A. S. (2023). *Statistical Rethinking with brms, ggplot2, and the tidyverse*. <https://bookdown.org/ajkurz/Statistical_Rethinking_recoded/>
- Bürkner, P.-C. (2017). brms: An R package for Bayesian multilevel models using Stan. *Journal of Statistical Software*, 80(1), 1–28.
- Gelman, A. & Hill, J. (2007). *Data Analysis Using Regression and Multilevel/Hierarchical Models*. Cambridge University Press.
- Bates, D., Mächler, M., Bolker, B., & Walker, S. (2015). Fitting linear mixed-effects models using lme4. *Journal of Statistical Software*, 67(1), 1–48.
:::
