---
title: "TP1: Intro to INLA(bru)"
subtitle: "Hierarchical models"
author: "Lionel Hertzog"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    number-sections: true
    code-fold: false
    code-copy: true
    code-overflow: wrap
    highlight-style: github
    theme: flatly
    fontsize: 1.05em
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
  eval: false
---

::: {.callout-note icon="false"}
## Objective

For this TP we will use a dataset containing 270 bird surveys made in 54 different sites within 8 regions. Each site was observed 5 times by different observers and every survey contains additionnal information including habitat type, elevation, canopy cover and temperature on the survey day.

The aim is to develop and test different mixed-effect models to explain the impact od the environment on bird counts.
:::

## Metadata

```{r eval=TRUE}
bird <- read.csv("bird_data.csv")
head(bird)
```

```{r eval=TRUE}
str(bird)
```


The columns are:

-  survey_id: an ID for the survey ranging from 1 to 270
-  region: an ID for the region of study ranging from 1 to 8
-  site: a site ID ranging from 1 to 54
-  observer: a character of the observer that did the survey with 6 levels
-  habitat: a character specifying the habitat type with three levels: forest, grassland and wetland
-  elevation_m: a numeric giving the elevation of the site in m 
-  canopy_pct: a numeric, canopy cover in percentage 
-  temp_C: the temperature the day of the survey in °C 
-  count: the number of birds observed


## Data Exploration

### Descriptice stats 

Explore the data, looking at descriptive statistics of the various variables: mean, sd, range, sample size per levels.

::: {.callout-tip icon="false" collapse="true"}
#### Hint
Check functions like `mean`, `sd`, `table`, `xtabs`. You an get the help of a R function by typing `?` in front of it, for instance:
```{r}
?mean
```
:::

### Correlations

Look at the correlation between the continuous variables including *count*

::: {.callout-tip icon="false" collapse="true"}
#### Hint

Check the function `cor` or `cor.test`.

:::


### Boxplots

Make boxplots of bird counts against (i) region and (ii) observer to explore variability in counts across these variables.

::: {.callout-tip icon="false" collapse="true"}
#### Hint

Check the function `boxplot` or use ggplot if already familiar with it.

:::

## Model fitting

### 

Fit a first model in inlabru with as response the bird count, as covariates habitat, elevation, canopy and temperature and a random intercept for observer.

Choose an appropriate distribution.

Take care 1: variables passed as random effects should be as integer (ID), so `observer` needs to be turned into `observer_id`

Take care 2: INLA(bru) does not automatically map factor variables to contrasts, you can use in the model formula:
```{r eval=FALSE}
habitat(main = habitat, model = "factor_contrast")
```

::: {.callout-tip icon="false" collapse="true"}
#### Hint 1

Remember that random intercept can be set using the `iid` model in INLA(bru):

```{r eval=FALSE}
f(observer_id, model = "iid")
```

:::

::: {.callout-tip icon="false" collapse="true"}
#### Hint 2

Here is an option to fit the model:

```{r eval=FALSE}
bird$observer_id <- as.integer(factor(bird$observer))
m1 <- bru(count ~ Intercept(1) + elevation_m + canopy_pct + temp_C + obs_eff(observer_id, model = "iid") + habitat(main = habitat, model = "factor_contrast"),  data = bird, family = "poisson")
```

:::

Explore the model output using `summary`, ``model$summary.random` ...


###

Add weakly informative PC prior to the random effect setting:

$$
P(\sigma > 4) = 0.01
$$
Also check other settings or if interested other prior distribution like `gamma` or `logtnormal`.

::: {.callout-tip icon="false" collapse="true"}
#### Hint

Setting priors on model components is done via the `hyper` argument:

```{r eval=FALSE}
obs_eff(observer_id, model = "iid", hyper = list(prec = list(prior = "pc.prec", param = c(4, 0.01))))
```

:::

Did this affect the estimated observer effect? Did the other model parameters changed?

###

Now we want to introduce a random site effect nested within a region effect.

To do this in INLA(bru), we pass two separate random effects one for region and one for site ensuring that site IDs are unique and not repeated.

Add also PC prior on these components.

Interpret and check the output, fixed effects, random terms ...

::: {.callout-tip icon="false" collapse="true"}
#### Hint

Here is how it can be done:

```{r eval=FALSE}
m2 <- bru(count ~ Intercept(1) + elevation_m + canopy_pct + temp_C + obs_eff(observer_id, model = "iid", hyper = list(prec = list(prior = "pc.prec", param = c(4, 0.01))))+ region_eff(region, model = "iid", hyper = list(prec = list(prior = "pc.prec", param = c(4, 0.01))))+ site_eff(site, model = "iid", hyper = list(prec = list(prior = "pc.prec", param = c(4, 0.01)))) + habitat(main = habitat, model = "factor_contrast"),  data = bird, family = "poisson")
```

:::

## Model checks

###

A hepful package to check model assumptions is `DHARMa`.

With INLa(bru) model we can use it as such:

1.  Generate model simulations matrix (rows observations, columns simulations) using the `generate` function, the generate functions should output simulated observations so $Poisson(e^{\eta_i})$ for this model.
2.  Pass these simulations to the `createDHARMa` function

::: {.callout-tip icon="false" collapse="true"}
#### Hint

Here is how it can be done:

```{r eval=FALSE, message=FALSE}
sims <- generate(m1, bird, ~ rpois(n = nrow(bird),
                                   lambda = exp(Intercept + elevation_m + canopy_pct + temp_C + habitat + obs_eff)))
ch_m1 <- createDHARMa(sims, bird$count)
```

:::

Do these steps for one of the two fitted model.

### 

Once a DHARMa object is created we can check various model assumptions.

Perform the following checks on the model:
-  overdispersion via `testDispersion`
-  outliers with `testOutliers`
-  residual distribution wit `plotResiduals`

What do you think? Do you see potential improvment on this model?

::: {.callout-tip icon="false" collapse="true"}
#### Hint

Given the evidence of overdispersion, a negative binomial is a natural next choice to improve the model.

```{r message=FALSE}
m3 <- bru(count ~ Intercept(1) + elevation_m + canopy_pct + temp_C + obs_eff(observer_id, model = "iid", hyper = list(prec = list(prior = "pc.prec", param = c(4, 0.01))))+ region_eff(region, model = "iid", hyper = list(prec = list(prior = "pc.prec", param = c(4, 0.01))))+ site_eff(site, model = "iid", hyper = list(prec = list(prior = "pc.prec", param = c(4, 0.01)))) + habitat(main = habitat, model = "factor_contrast"),  data = bird, family = "nbinomial")
```

:::

###

Explore various options to improve the model fit and try to achieve good model checks.

::: {.callout-tip icon="false" collapse="true"}
#### Hint

Check various alternative family `nbinomial` potentially with variants as defined in `inla.doc("nbinomial")`.

:::


## Model interpretation

###

We want know to plot the estimated effects of temperature and canopy cover on bird count.

Derive model predictions along gradients of temperature and canopy cover, and plot the resulting regression lines together with their confidence bands.

::: {.callout-tip icon="false" collapse="true"}
#### Hint

Create first a new data frame with a gradient in temperature while all other variables are held constant to their mean:

```{r}
newdat_temp <- data.frame(temp_C = seq(min(bird$temp_C),
                                       max(bird$temp_C),
                                       length.out = 10),
                          elevation_m = mean(bird$elevation_m),
                          canopy_pct = mean(bird$canopy_pct))

pred_temp <- predict(m3, newdat_temp, ~ exp(Intercept + elevation_m + canopy_pct + temp_C))

ggplot(pred_temp, aes(x=temp_C, y=mean, ymin=q0.025, ymax=q0.975)) +
  geom_ribbon(alpha = 0.1) +
  geom_path() +
  theme_classic() +
  labs(x = "Temperature (°C)",
       y = "Predicted bird counts (with 95% CI")

```

:::

###

Now we want to explore the variations in bird counts between the site within the regions and between the observers.

We can access the estimated deviation from the average bird counts per levels of the random terms via `m$summary.random` which is a list of dataframes with the estimated random terms.

Plot first the estimated observer deviations. What do you think?

Plot then the site within regions deviations, to do this add the region random effects to the adequate site random effects.

::: {.callout-tip icon="false" collapse="true"}
#### Hint

For the site within region deviations we can use the `predict` function.

```{r}
pred_site <- predict(
  m3,
  newdata = unique(bird[, c("site", "region")]),
  formula = ~ region_eff + site_eff
)

ggplot(pred_site, aes(x=fct_reorder(factor(site), mean),
                      y = mean,
                      ymin=q0.025,
                      ymax=q0.975)) +
  geom_linerange() +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  labs(x = "Site",
       y = "Deviation from average")

```

:::
